---
title: "NCBI SRA submission"
author: "Sarah Tanja"
date: 08/19/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "" # Prevents appending '##' to beginning of lines in code output
)        
```

```{r library}
library(tidyverse)
```

# 1 Study Background

Corinne Klohmann (CK) collected surface eDNA water samples via kayak
from three sites across Puget Sound: Golden Gardens, Smith's Cove, and
Alki Beach. At each site, surface water samples were collected over both
a 100m transect of eelgrass (Zostera marina), and a 100m transect of
non-eelgrass habitat at 1, 20, 40, 60, 80, and 100m along each transect.
The aim of this analyses is to compare bacterial community of the
seawater over eelgrass with non-eelgrass samples. The hypothesis is that
eelgrass will filter out pathogenic bacteria from the water column, and
that we will detect less pathogenic bacteria in samples taken from the
eelgrass-associated transects. This study was conducted by CK in 2020.

# 2 NCBI Submission

### Sequence Read Archive submission SUB16423997

#### Submission title:

> Seasonal changes in pathogenic bacteria around eelgrass (Zostera
> marina) meadows in Puget Sound, Aug 20 \`26

#### Sample type:

> MIMARKS Survey related, water - Use for any type of marker gene
> sequences, eg, 16S, obtained directly fromt he environment, without
> culturing or identification of the organisms.

<<<<<<< HEAD
# 3 Biosample required attributes
=======
# 3 Required attributes
>>>>>>> 57eca64 (metadata clean)

These are the metadata columns that are required for NCBI submission!
For this study we follow the BioSample Attributes for
`package MIMARKS: survey, water, version 6.0` .

#### sample_name

- Our sample names follow the format `mm-dd-YY-[gg/alki/sc]-[g/ng][1-5]`
  , where:

  - mm-dd-YY are the sample collection dates

  - gg = Golden Gardens, alki = Alki Beach, sc = Smith's Cove

  - g = grass, ng = no grass

  - 1 = 1m from shore, 2 = 20m from shore, 3 = 40m from shore, 4 = 60m
    from shore, 5 = 80m from shore

#### organism

- seawater metagenome
  - These samples were seawater eDNA samples, using the 16S rRNA marker
    gene for bacterial identification

> (see [Organism
> information](https://www.ncbi.nlm.nih.gov/biosample/docs/organism/#:~:text=2%20ABC%2D2021-,Metagenomes%20(Microbiomes),-We%20consider%20any)
> for why we chose this)

#### collection_date

> The date on which the sample was collected; date/time ranges are
> supported by providing two dates from among the supported value
> formats, delimited by a forward-slash character; collection times are
> supported by adding "T", then the hour and minute after the date, and
> must be in Coordinated Universal Time (UTC), otherwise known as "Zulu
> Time" (Z); supported formats include "DD-Mmm-YYYY", "Mmm-YYYY", "YYYY"
> or ISO 8601 standard "YYYY-mm-dd", "YYYY-mm", "YYYY-mm-ddThh:mm:ss";
> e.g., 30-Oct-1990, Oct-1990, 1990, 1990-10-30, 1990-10,
> 21-Oct-1952/15-Feb-1953, 2015-10-11T17:53:03Z; valid non-ISO dates
> will be automatically transformed to ISO format

#### env_broad_scale

- estuarine biome

> Add terms that identify the major environment type(s) where your
> sample was collected. Recommend subclasses of biome \[ENVO:00000428\].
> Multiple terms can be separated by one or more pipes e.g.:  mangrove
> biome \[ENVO:01000181\]\|estuarine biome \[ENVO:01000020\]

#### env_local_scale

- shoreline\|eelgrass meadow

- shoreline\|bare substrate

> Add terms that identify environmental entities having causal
> influences upon the entity at time of sampling, multiple terms can be
> separated by pipes, e.g.:  shoreline \[ENVO:00000486\]\|intertidal
> zone \[ENVO:00000316\]

#### env_medium

- estuarine water

> Add terms that identify the material displaced by the entity at time
> of sampling. Recommend subclasses of environmental material
> \[ENVO:00010483\]. Multiple terms can be separated by pipes e.g.:
> estuarine water \[ENVO:01000301\]\|estuarine mud \[ENVO:00002160\]

#### geo_loc_name

- USA: Puget Sound, Golden Gardens

- USA: Puget Sound, Alki Beach

- USA: Puget Sound, Smith's Cove

> Geographical origin of the sample; use the appropriate name from this
> list
> <https://www.insdc.org/submitting-standards/geo_loc_name-qualifier-vocabulary/>.
> Use a colon to separate the country or ocean from more detailed
> information about the location, eg "Canada: Vancouver" or "Germany:
> halfway down Zugspitze, Alps"

#### lat_lon

- The geographical coordinates of the location where the sample was
  collected. Specify as degrees latitude and longitude in format
  "d\[d.dddd\] N\|S d\[dd.dddd\] W\|E", eg, 38.98 N 77.11 W

#### depth

- All of these samples were taken at the surface and therefore depth
  will be 0 for all

> Depth is defined as the vertical distance below surface, e.g. for
> sediment or soil samples depth is measured from sediment or soil
> surface, respectivly. Depth can be reported as an interval for
> subsurface samples.

# 4 Wrangle metadata

Let's checkout the metadata!

- slice(-1) removes the first data row after the headers, which in the
  `Metadata_cak.tsv` file was descriptions of the column data types.

```{r metadatapeek}
metadata <- read_tsv("../metadata/Metadata_cak.tsv") |> 
  slice(-1)

# make dist from shore a double not a character
metadata <- metadata |> 
  mutate(dist_from_shore = as.double(dist_from_shore))

# make date column
metadata <- metadata |> 
  mutate(
    date = stringr::str_extract(`sample-id`, "^\\d{1,2}-\\d{1,2}") |>
      paste0("-2020") |>
      as.Date(format = "%m-%d-%Y")
  )
# clean up month column
metadata <- metadata |> 
  rename(month_no = month)

knitr::kable(metadata, format="pipe")
```

Our metadata file has 6 columns:

- sample-id

- barcode-sequence

- seagrass

- month

- site

- dist_from_shore

::: callout-note
Q: The sample-id indicates the winter samples were collected December
2020... were the Summer samples collected earlier that same year in
August 2020, or later in August 2021?

A: All samples were collected in the year 2020!
:::

We have 60 water samples across 3 sites (`golden_gardens`, `alki`,
`smiths_cove`) that were collected either in Winter (12 / December on
12/16, 12/18, and 12/22 ) or in Summer (8 / August on 8/25 and 8/21),
over eelgrass habitat transects (`grass`) or over sediment habitat
transects (`none`), at 1, 20, 40, 60, and 80 meters from shore along a
100m transect.

```{r metadatacount}
metadata |> count(site, month_no, seagrass, dist_from_shore)
```

We don't have any repeat samples across all of our metadata columns, the
distance from shore column could be collapsed and interpreted as our
repeated sampling per site, transect, and date.

```{r metadatacount2}
metadata |> count(site, month_no, seagrass)
```

When we group all the samples from the same day and place (and collapse
the dist-from_shore column ), we have a sample size of 5.

The following fielddata are environmental conditions recorded during
sampling for Enterococcus that coincided with sampling for eDNA. The
Enterococcus samples were taken in triplicate, while the eDNA samples
were single samples.

The Enterococcus samples were also taken in February, March, June, July,
August, October, November, December of 2020 and in January of 2021 -
while the eDNA samples we have are only from August and December 2020.

```{r fieldinspect}
fielddata <- read_csv("../metadata/FieldAssayCounts.csv")

# Filter ton metadata only from Aug and Dec and format as.date
fielddata <- fielddata |> 
  filter(month == c("August", "December")) |> 
  mutate(
    ent_date = as.Date(Date, format = "%m/%d/%Y")
  )

unique(fielddata$month)
unique(fielddata$ent_date)
```

```{r fieldclean}

fieldclean <- fielddata |> 
  mutate(
    month_no = case_when(
      month == "August" ~ 8,
      month == "December" ~ 12, 
      TRUE ~ NA_real_
    ), 
    site = case_when(
      site == "Alki" ~ "alki", 
      site == "smithcove" ~ "smiths_cove",
      site == "GG" ~ "golden_gardens",
      site == "Disc" ~ "disc"
    ), 
    seagrass = case_when(
    grass.nograss == 1 ~ "grass",
    grass.nograss == 0 ~ "none"
  )) |> 
  rename(
      dist_from_shore = dist.from.shore
    ) |> 
  select(ent_date, month, month_no, site, transect, seagrass, dist_from_shore, temp, ph, sal, par, do, ent_count
    )

fieldclean  
```

The field assay counts are data from the enterococcus water samples that
were collected in the same location and time as the eDNA water samples.
The enterococcus water samples were collected in triplicate, as were
their metadata, so we have to average that metadata to get 1 row of
information for each eDNA water sample.

```{r fieldaverages}
fieldavg <- fieldclean |> 
  group_by(ent_date, site, seagrass, dist_from_shore) |> 
  summarize(
    n        = n(), # counts the input rows, should be 3
    avg_temp = mean(temp, na.rm = TRUE),
    avg_ph   = mean(ph, na.rm = TRUE),
    avg_sal  = mean(sal, na.rm = TRUE),
    avg_par  = mean(par, na.rm = TRUE),
    avg_do   = mean(do, na.rm = TRUE),
    avg_ent  = mean(ent_count, na.rm = TRUE),
    .groups = "drop"
  )

fieldavg
```

`fieldavg` should have an average metadata row for each sampling
location (site x seagrass x dist_from_shore)

- The `fieldavg$n` column should always be 3... indicating that three
  samples from the same location were averaged

- I notice some of them are 4 or 2 instead...

- Let's check that by filtering for any that are not 3...

```{r}
fieldavg |> 
  filter(n != 3)
```

::: callout-note
OK... 6 of our samples don't group into the triplicate sample
structure... They're all in the dist_from_shore 1 sampling location...
And it looks like there is a sample labelled `grass` that should have
been `none`... because the grass category has an n of 4 while the none
category has an n of 2. Let's explore that...
:::

When I look at `fielddata` I found that Transects A, B, C should all be
over seagrass, while transects D, E, F should all be over sediment. It
seems like the first entry of Transect D is mislabelled with a 1,
indicating a seagrass sample. It should be a 0. Let's fix it:

```{r}
fieldclean <- fieldclean |>
  mutate(
    seagrass = case_when(
      transect == "D" ~ "none",
      TRUE ~ seagrass
    )
  )
```

Now let's re-run the averaging step and search for any n that are not 3

```{r}
fieldavg <- fieldclean |> 
  group_by(ent_date, site, seagrass, dist_from_shore) |> 
  summarize(
    n        = n(), # counts the input rows, should be 3
    avg_temp = mean(temp, na.rm = TRUE),
    avg_ph   = mean(ph, na.rm = TRUE),
    avg_sal  = mean(sal, na.rm = TRUE),
    avg_par  = mean(par, na.rm = TRUE),
    avg_do   = mean(do, na.rm = TRUE),
    avg_ent  = mean(ent_count, na.rm = TRUE),
    .groups = "drop"
  )

fieldavg |> filter(n !=3)
fieldavg
```

::: callout-important
Ok, that appears to have fixed it! IMPORTANT NOTE, this should be fixed
for the Enterococcus linear models as well if that metadata is the same
in the other analysis...
:::

::: callout-warning
The `fieldavg` dataframe only has 36 rows... and I expected 60 (one for
each eDNA sample) ... I can see that the issue is dist_from_shore in
`fieldavg` only records 1, 40, 80 in Aug and 20, 60, 100 in December ,
meanwhile `metadata` for our eDNA samples have 1, 20, 60, 80 for each
transect.
:::

The sample collection date from the enterococcus metadata for December
golden gardens is the 21st, but it's the 22nd in the eDNA metadata...

::: callout-note
- Q: Is this a typo or were samples collected on separate days?
:::

```{r metadatajoin}
metadatajoin <- metadata |> 
  left_join(fieldsummary, by = join_by(site, seagrass, dist_from_shore, closest(x$date <= y$ent_date))) |> 
  select(!ent_date)
```

# 5 Inspect fastq.gz files

```{r fastqlist, engine='bash'}
ls ../fastq_files
```

```{r fastqcount, engine='bash'}
ls -1 ../fastq_files/*fastq.gz* | wc -l
```

We have 118 fastq.gz files... There should be a R1 and an R2 for each
sample.

```{r div2}
118/2
```

So we have fastq files for 59 out of the 60 samples...

# 6 Generate checksums

Generate md5 for each fastq.gz and save it in the data directory

```{r md5sum, engine='bash'}

for file in ../fastq_files/*.fastq.gz; do
    md5sum "$file" > "${file}.md5"
done
```

We need to match our filenames with our metadata.

Each filename follows the pattern `sample-id_S##_L001_R1/2.fastq.gz` and
in metadata the sample-id coumn should match the sample-id prior to the
\_S##\_L001_R1/R2.fastq.gz

I'd like to add a R1 column and a R2 column and put the filepaths to the
forward and reverse reads into the metadata dataframe. This is the
format that we will need for the NCBI submission.

```{r fastqlookup}

fastq_lookup <- tibble(
  filepath = list.files(
    "../fastq_files",
    pattern = "\\.fastq\\.gz$",
    full.names = TRUE
  )
) |>
  mutate(
    filename = basename(filepath),
    `sample-id` = str_extract(filename, "^[^_]+"),
    read = str_extract(filename, "(?<=_)R[12](?=_)"),
    
    # checksum path needed temporarily to read the file
    checksum_path = paste0(filepath, ".md5"),
    
    # checksum filename
    checksum_filename = basename(checksum_path),
    
    # checksum value
    checksum = map_chr(
      checksum_path,
      \(x) readLines(x, n = 1) |>
        str_extract("^[[:xdigit:]]+")
    )
  ) |>
  select(
    `sample-id`,
    read,
    filename,
    checksum_filename,
    checksum
  ) |>
  pivot_wider(
    id_cols = `sample-id`,
    names_from = read,
    values_from = c(filename, checksum_filename, checksum),
    names_glue = "{read}_{.value}"
  )
```

Now we have `fastq_lookup` which has the sample-id, the fastq filenames
to the R1 and the R2 `.fastq.gz` files, ad the names of the md5 checksum
files and the checksum values for each R1 and R2 file.

# 7 Join metadata

We need to add the required NCBI attributes to our metadata file.

```{r addNCBIattributes}
ncbi-metadata <- metadatajoin |> 
  mutate(
    sample_name = sample-id, 
    organism = `water metagenome`,
    collection_date = date,
    env_broad_scale = ``,
    env_local_scale = ``,
    env_medium = ``,
    geo_loc_name = ``,
    lat_lon = ``,
    depth = 0
  )
```

# Join fastq_lookup to metadata

We can join these filepaths to the metadata and look for which sample-id
is lacking fastq files!

```{r metadatajoin}
metadata <- metadata |>
  left_join(fastq_lookup, by = "sample-id")

metadata |>
  filter(is.na(R1) | is.na(R2))
```

::: callout-note
OK so sample `8-25-sc-ng3` is missing fastq.gz files. If we can't locate
the fastq files for this sample we will exclude this sample from the
analysis.
:::

# 8 FTP File Upload
